---
title: "Chicago Main Analyses"
author: "Renata Poulton Kamakura"
date: "2022-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gjam)

##Group means for segment data
library(dplyr)

##Visuals
library(ggplot2)

#install.packages( "./ordReg_1.0.tar.gz", repos=NULL, type="source" )
#library('ordReg')

```

```{r read in tree lvl data}
##read in data
TreeLvlCombined <- read.csv("../01Data/Chicago/ChicagoAllYearsComb.csv")
```


##Function to get sigma
```{r sum squares}
sumSquares <- function(predicted, actual){
  
  ss <- 0
  
  for(s in 1:length(predicted)){
    
    ss <- ss + (predicted[s] - actual[s])^2
    
    if( is.na(ss)){
      
      print(paste("ERROR: at iteration", s, "ss became NA"))
      
      break
      
    }
    
  }
  
  return(ss)
}

basalArea <- function(dbh){
  dbh_cm <- dbh*2.54
  ba <- (pi * (dbh_cm/2)^2)/10000
  return(ba)
  
}

```

##Multiple Tree Health Variables

```{r pull in tree-lvl data}
#tree species information
SppSpecificData <- read.csv("../01Data/Chicago/SpeciesInformation_Chicago_Dirr.csv")

#for ease later
TreeLvlCombined$GenSpe <- paste(TreeLvlCombined$Genus, TreeLvlCombined$Species, sep="_")

#To line up with species info
TreeLvlCombined$Species[TreeLvlCombined$Species == "unknown"] <- "spp."

TreeLvlCombined$TreeSize <- rep(NA, length(TreeLvlCombined$TreeName))
TreeLvlCombined$UrbanApp <- rep(NA, length(TreeLvlCombined$TreeName))

names(TreeLvlCombined)[names(TreeLvlCombined) == "CrownVigor"] <- "CrownStress"

##for a few weird species
londonplane <- c("x acerifolia", "hybrida")
TreeLvlCombined$Species[TreeLvlCombined$Species %in% londonplane] <- "x hispanica"

TreeLvlCombined$Species[TreeLvlCombined$Species == "sylvatica"] <- "silvatica"

#row by row
for (t in 1:length(TreeLvlCombined$X)){
  YesGen <- SppSpecificData$SUB_GENUS == TreeLvlCombined$Genus[t]
  YesSpp <- SppSpecificData$SUB_SPECIES == TreeLvlCombined$Species[t]
  
  SppInfoInd <- c(1:length(SppSpecificData[,1]))[(YesGen & YesSpp)]
  
  #what do you do if you don't have info on this species
  if(length(SppInfoInd) == 0){
    print(paste("Error, no species information for", TreeLvlCombined$GenSpe[t], "Tree #", t))
    
    ##try to fill it in with the default for the genus
    Spp2 <- SppSpecificData$SUB_SPECIES =="spp."
    SppInfoInd2 <- c(1:length(SppSpecificData[,1]))[(YesGen & Spp2)]
    
    if(length(SppInfoInd2) == 0){
      print(paste("Error, no generic genus-level information for", TreeLvlCombined$Genus[t], "Tree #", t))
      next #go to the next tree
    } else {
      TreeLvlCombined$TreeSize[t] <-  mean(c(SppSpecificData$Height_height[SppInfoInd2], SppSpecificData$Height_low[SppInfoInd2]))
      TreeLvlCombined$UrbanApp[t] <- SppSpecificData$StreetUrban[SppInfoInd2]
    }
  
  #have a species-level match
  } else {
    TreeLvlCombined$TreeSize[t] <-  mean(c(SppSpecificData$Height_height[SppInfoInd], SppSpecificData$Height_low[SppInfoInd]))
    TreeLvlCombined$UrbanApp[t] <- SppSpecificData$StreetUrban[SppInfoInd]
  } #seeing if there is a match
  
}

#Prep for analyses
TreeLvlCombined$UrbanApp <- as.factor(TreeLvlCombined$UrbanApp)

```


## Grouping for sample sizes

Goal here is to do some simplification because we don't have the sample size to characterize all the possible predictor levels or even all of the response levels

```{r prep response data}
#in case there are errors, easy to go back
TreeLvlCombined_mod <- TreeLvlCombined

#Get rid of trees without local site conditions
TreeLvlCombined_mod <- TreeLvlCombined_mod[!is.na(TreeLvlCombined_mod$Gardenscape),] #eliminates 26 obs

#defoliation
TreeLvlCombined_mod$Defoliation[TreeLvlCombined_mod$Defoliation > 3] <- 3
TreeLvlCombined_mod$Defoliation <- round(TreeLvlCombined_mod$Defoliation) #Get rid of 0.5

#discoloration
TreeLvlCombined_mod$Discoloration[TreeLvlCombined_mod$Discoloration > 4] <- 4
TreeLvlCombined_mod$Discoloration <- round(TreeLvlCombined_mod$Discoloration) #Get rid of 0.5

#dieback
TreeLvlCombined_mod$Dieback[((TreeLvlCombined_mod$Dieback > 3) & (TreeLvlCombined_mod$Dieback <= 6))] <- 4 #11-25%
TreeLvlCombined_mod$Dieback[(TreeLvlCombined_mod$Dieback > 6)] <- 5 #25%+
TreeLvlCombined_mod$Dieback[TreeLvlCombined_mod$CrownStress >= 5] <- 5 #trees are dead
TreeLvlCombined_mod$Dieback <- round(TreeLvlCombined_mod$Dieback) #Get rid of 0.5

#crown stress
TreeLvlCombined_mod$CrownStress <- round(TreeLvlCombined_mod$CrownStress, digits=0)
```

```{r prep baseline and tree characteristics}

#####baseline/sampling variables
#prep segment
TreeLvlCombined_mod$Segment <- as.character(TreeLvlCombined_mod$Segment)
TreeLvlCombined_mod<- TreeLvlCombined_mod[TreeLvlCombined_mod$Segment != "UIC",] #not include
TreeLvlCombined_mod$Segment[TreeLvlCombined_mod$Segment == "324"] <- "01_324" #set as baseline
TreeLvlCombined_mod$Segment <- as.factor(TreeLvlCombined_mod$Segment)

#Year
TreeLvlCombined_mod$Year <- as.factor(TreeLvlCombined_mod$Year)

############Tree characteristics
##prep basal area
TreeLvlCombined_mod$BasalArea <- sapply(TreeLvlCombined_mod$TotalDBH, basalArea)
TreeLvlCombined_mod <- TreeLvlCombined_mod[!is.na(TreeLvlCombined_mod$BasalArea),]

TreeLvlCombined_mod <- TreeLvlCombined_mod[!(TreeLvlCombined_mod$GenSpe == "_unknown"),]
#group the tree size variable
thresh1 <- 35 #should check if this is a reasonable grouping
thresh2 <- 50
ts_initial <- TreeLvlCombined_mod$TreeSize
TreeLvlCombined_mod$TreeSize[ts_initial <= thresh1] <- "Small"
TreeLvlCombined_mod$TreeSize[ts_initial > thresh1 & ts_initial <= thresh2] <- "Medium"
TreeLvlCombined_mod$TreeSize[ts_initial > thresh2] <- "Large"

TreeLvlCombined_mod$TreeSize  <- as.factor( TreeLvlCombined_mod$TreeSize )

#urban appropriate
TreeLvlCombined_mod$UrbanApp <- as.factor(TreeLvlCombined_mod$UrbanApp)
```

```{r prep site characteristics}
################ Site characteristics
#######Land use
TreeLvlCombined_mod$LandUse <- as.character(TreeLvlCombined_mod$LandUse) #easier to modify
##leaving the same:multi-family residential (MFR), institutional (INST), managed park (MP)
###residential - set as baseline
TreeLvlCombined_mod$LandUse[TreeLvlCombined_mod$LandUse == "SFR-A" | TreeLvlCombined_mod$LandUse == "SFR-D"] <- "AA_SFR" #collapse into single-family res
##Commercial + Mixed
TreeLvlCombined_mod$LandUse[TreeLvlCombined_mod$LandUse == "COMM" | TreeLvlCombined_mod$LandUse == "MIX"] <- "COMM_Mx"
#Other (industrial + vacant)
TreeLvlCombined_mod$LandUse[TreeLvlCombined_mod$LandUse == ""| TreeLvlCombined_mod$LandUse == "V" | TreeLvlCombined_mod$LandUse == "UT" | TreeLvlCombined_mod$LandUse == "IND" | TreeLvlCombined_mod$LandUse == "TR" | TreeLvlCombined_mod$LandUse == "AG" ] <- "Other"
#turn back into factor
TreeLvlCombined_mod$LandUse <- as.factor(TreeLvlCombined_mod$LandUse)


#####Percent Impervious
TreeLvlCombined_mod$PercentImperv[TreeLvlCombined_mod$PercentImperv <=2] <- 1 #less than 26% imperv
TreeLvlCombined_mod$PercentImperv[TreeLvlCombined_mod$PercentImperv ==3] <- 2 #26-50
TreeLvlCombined_mod$PercentImperv[TreeLvlCombined_mod$PercentImperv ==4] <- 3 #51-75
TreeLvlCombined_mod$PercentImperv[TreeLvlCombined_mod$PercentImperv ==5] <- 4 #76-100

TreeLvlCombined_mod <- TreeLvlCombined_mod[!is.na(TreeLvlCombined_mod$PercentImperv),] #gets rid of 35, 26 of which are also removed with gardenscape
TreeLvlCombined_mod$PercentImperv <- as.factor(TreeLvlCombined_mod$PercentImperv)

######Site type
TreeLvlCombined_mod$SiteType <- as.character(TreeLvlCombined_mod$SiteType) #to edit more easily

TreeLvlCombined_mod$SiteType[is.null(TreeLvlCombined_mod$SiteType) |TreeLvlCombined_mod$SiteType== ""| TreeLvlCombined_mod$SiteType== "MP" | TreeLvlCombined_mod$SiteType == "OM"] <- "Other"

TreeLvlCombined_mod$SiteType[TreeLvlCombined_mod$SiteType == "SC"] <- "AA_SC" #set as baseline

TreeLvlCombined_mod$SiteType <- as.factor(TreeLvlCombined_mod$SiteType)

########Crown Light
TreeLvlCombined_mod$CrownLight <- as.numeric(TreeLvlCombined_mod$CrownLight)
TreeLvlCombined_mod$CrownLight <- round(TreeLvlCombined_mod$CrownLight)
TreeLvlCombined_mod$CrownLight[TreeLvlCombined_mod$CrownLight == 0] <- 1
#TreeLvlCombined_mod$CrownLight[is.na(TreeLvlCombined_mod$CrownLight)] <- 3 #just give it a moderate value
TreeLvlCombined_mod$CrownLight <- as.factor(TreeLvlCombined_mod$CrownLight)

#####powerlines
TreeLvlCombined_mod$Powerline[TreeLvlCombined_mod$Powerline] <- 1
TreeLvlCombined_mod$Powerline[TreeLvlCombined_mod$Powerline == FALSE] <- 0

```

```{r prep cues to care}

#Translate Gardenscape into text
TreeLvlCombined_mod$Gardenscape[TreeLvlCombined_mod$Gardenscape==""] <- "None"
h1 <- c("None", "Minimal", "Moderate", "Extensive")
h2 <- c(1:4)
GDict <- as.data.frame(matrix(c(h1,h2), nrow=4, ncol=2, byrow=FALSE))
names(GDict) <- c("Original", "Simplified")

TreeLvlCombined_mod$Gardenscape <- GDict$Simplified[match(TreeLvlCombined_mod$Gardenscape, GDict$Original)]

##gardenscape
TreeLvlCombined_mod$Gardenscape <- as.character(TreeLvlCombined_mod$Gardenscape)
TreeLvlCombined_mod$Gardenscape[TreeLvlCombined_mod$Gardenscape == 3] <- 4
TreeLvlCombined_mod$Gardenscape[TreeLvlCombined_mod$Gardenscape == 2] <- 3
TreeLvlCombined_mod$Gardenscape[TreeLvlCombined_mod$Gardenscape == 1] <- 2
TreeLvlCombined_mod$Gardenscape[TreeLvlCombined_mod$Gardenscape == 2 & TreeLvlCombined_mod$Mowing %in% c("old", "potentially never")] <- 1
TreeLvlCombined_mod$Gardenscape <- as.factor(TreeLvlCombined_mod$Gardenscape)

#Pruning
TreeLvlCombined_mod$PrunedCorrect <-TreeLvlCombined_mod$Pruned
TreeLvlCombined_mod$PrunedCorrect[TreeLvlCombined_mod$HatrackPrune == 1 | TreeLvlCombined_mod$FlushCutPrune == 1] <- 0
TreeLvlCombined_mod$PrunedCorrect <- as.factor(TreeLvlCombined_mod$PrunedCorrect )

##create Mulch categories
TreeLvlCombined_mod$Mulch[TreeLvlCombined_mod$Mulch == 1] <- "Mulch"
TreeLvlCombined_mod$Mulch[TreeLvlCombined_mod$Mulch == 0] <- "AA_None"
TreeLvlCombined_mod$Mulch[TreeLvlCombined_mod$MulchVolcano == 1] <- "Volcano"

TreeLvlCombined_mod$Mulch <- as.factor(TreeLvlCombined_mod$Mulch)


# ##create a correctly mulched metric in place
# TreeLvlCombined_mod$CorrMulch <- TreeLvlCombined_mod$Mulch
# TreeLvlCombined_mod$CorrMulch <- as.character(TreeLvlCombined_mod$CorrMulch)
# TreeLvlCombined_mod$CorrMulch[TreeLvlCombined_mod$CorrMulch == "Mulch"] <- 1
# TreeLvlCombined_mod$CorrMulch[TreeLvlCombined_mod$CorrMulch != 1] <- 0
# TreeLvlCombined_mod$CorrMulch <- as.factor(TreeLvlCombined_mod$CorrMulch)
```

```{r prep species}
#############Species group adjustment

#fix spelling errors and collapse uncertain IDs
TreeLvlCombined_mod$GenSpe[TreeLvlCombined_mod$Genus == "Fraxinus"] <- "Fraxinus spp."
TreeLvlCombined_mod$GenSpe[(TreeLvlCombined_mod$GenSpe == "Tilia_cordata") | (TreeLvlCombined_mod$GenSpe == "Tilia_americana") | TreeLvlCombined_mod$GenSpe == "Tilia_tomentosa"] <- "Tilia spp." #small sample size and some cultivars
TreeLvlCombined_mod$GenSpe[TreeLvlCombined_mod$GenSpe == "Ulmus_americana"] <- "Ulmus spp." #unsure IDs
TreeLvlCombined_mod$GenSpe[TreeLvlCombined_mod$GenSpe == "Ulmus_pumila"| TreeLvlCombined_mod$GenSpe == "Ulmus_unknown" | TreeLvlCombined_mod$GenSpe == "Ulmus_spp."] <- "Ulmus spp." #unsure IDs

#get a list of species with at least 50 observations
SppFreq <- sort(table(TreeLvlCombined_mod$GenSpe),decreasing = TRUE)
CommonSpp <- SppFreq[SppFreq >= 72]
#confirm separately that have at least 35 individual trees
#look at genera frequency for the others
remainingSpp <- TreeLvlCombined_mod$Genus[!(TreeLvlCombined_mod$GenSpe %in% names(CommonSpp))]
GenusFreq <- table(remainingSpp)
CommonGenus <- GenusFreq[GenusFreq > 72]


#Create a species group category
#leave the common species alone: G. triacanthos, Ulmus, A. platanoides, Fraxinus spp., Tilia spp., C. occidentalis, G. dioicus
#then you have these genera with enough spp to work with
#group the rest by size
#going to do this backwards to make it easier to implement

TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$TreeSize == "Small"] <- "Other_sm" #group by size
TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$TreeSize == "Medium"] <- "Other_md" #group by size
TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$TreeSize == "Large"] <- "Other_lg" #group by size
TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$Genus %in% names(CommonGenus)]<- TreeLvlCombined_mod$Genus[TreeLvlCombined_mod$Genus %in% names(CommonGenus)] #common genera
TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$GenSpe %in% names(CommonSpp)] <- TreeLvlCombined_mod$GenSpe[TreeLvlCombined_mod$GenSpe %in% names(CommonSpp)] #common species
#choose baseline species - most common species we saw: Gleditsia triacanthos
TreeLvlCombined_mod$SppGroup[TreeLvlCombined_mod$SppGroup == "Gleditsia_triacanthos"] <- "AA_Gleditsia_triacanthos"
TreeLvlCombined_mod <- TreeLvlCombined_mod[!is.na(TreeLvlCombined_mod$SppGroup),] #only gets rid of one
TreeLvlCombined_mod$SppGroup <- as.factor(TreeLvlCombined_mod$SppGroup) #for model

```

## Check correlations

Need to update this still
```{r between pred corr}
##Pull in relevant R packages
library(psych)####Tetrachoric - binary
#tetrachoric(data)

library(polycor)##Polychoric - ordinal
#polychor(x,y)

library(rcompanion)##Cramer's V - nominal
#cramerV(data)

#Look at the variables
preds <- c("TreeName", "Segment", "Year", "BasalArea", "SppGroup","Mulch", "PercentImperv", "Powerline", "UrbanApp", "SiteType", "Gardenscape","PrunedCorrect", "CrownLight", "LandUse")

#binary
binVars <- c("Powerline","UrbanApp","PrunedCorrect")
oVars <- binVars #so you can do pairwise comparisons

for(b in 1:length(binVars)){
  #pick starting variable
  b1 <- binVars[b]
  oVars <- oVars[oVars != b1] #the other ones left
  
  #if none left
  if(length(oVars) == 0){
    print("Done with all pairwise comparisons!")
    next
  } 
  
  #for each remaining combination
  for (c in 1:length(oVars)){
    b2 <- oVars[c]
    tb <- table(TreeLvlCombined_mod[,c(b1,b2)])
    
    tet <- tetrachoric(x=tb,)
    
    print(paste("The tetrachoric rho for", b1, "and", b2, "is:", tet$rho))
  }
  
}

# [1] "The tetrachoric rho for Powerline and UrbanApp is: 0.0419419222489716"
# [1] "The tetrachoric rho for Powerline and PrunedCorrect is: -0.173576701820483"
# [1] "The tetrachoric rho for UrbanApp and PrunedCorrect is: -0.0332825610940145"

#ordinal
ordVars <- c("PercentImperv","Gardenscape","CrownLight", "LandUse")
oVars_o <- ordVars #so you can do pairwise comparisons

for(o in 1:length(ordVars)){
  #pick starting variable
  o1 <- ordVars[o]
  oVars_o <- oVars_o[oVars_o != o1] #the other ones left
  
  #if none left
  if(length(oVars_o) == 0){
    print("Done with all pairwise comparisons!")
    next
  } 
  
  #for each remaining combination
  for (p in 1:length(oVars_o)){
    o2 <- oVars_o[p]
    
    pol <- polychor(TreeLvlCombined_mod[,c(o1)],TreeLvlCombined_mod[,c(o2)])
    print(paste("The polychoric correlation for", o1, "and", o2, "is:", pol))
  }
  
}

# [1] "The polychoric correlation for PercentImperv and Gardenscape is: 0.0312264641687608"
# [1] "The polychoric correlation for PercentImperv and CrownLight is: -0.115029018440944"
# [1] "The polychoric correlation for PercentImperv and LandUse is: 0.0244584005182064"
# [1] "The polychoric correlation for Gardenscape and CrownLight is: -0.111718549227959"
# [1] "The polychoric correlation for Gardenscape and LandUse is: -0.0983578365799617"
# [1] "The polychoric correlation for CrownLight and LandUse is: -0.0327372843075162"

#nominal
nomVars <- c("Segment", "SppGroup","Mulch","SiteType")
oVars_n <- nomVars #so you can do pairwise comparisons

for(n in 1:length(nomVars)){
  #pick starting variable
  n1 <- nomVars[n]
  oVars_n <- oVars_n[oVars_n != n1] #the other ones left
  
  #if none left
  if(length(oVars_n) == 0){
    print("Done with all pairwise comparisons!")
    next
  } 
  
  #for each remaining combination
  for (m in 1:length(oVars_n)){
    n2 <- oVars_n[m]
    tb_n <- table(TreeLvlCombined_mod[,c(n1,n2)])
    
    cramV <- cramerV(tb_n)
    
    print(paste("Cramer's V for", n1, "and", n2, "is:", cramV))
  }
  
}

# [1] "Cramer's V for Segment and SppGroup is: 0.2939"
# [1] "Cramer's V for Segment and Mulch is: 0.4172"
# [1] "Cramer's V for Segment and SiteType is: 0.3559"
# [1] "Cramer's V for SppGroup and Mulch is: 0.1726"
# [1] "Cramer's V for SppGroup and SiteType is: 0.1682"
# [1] "Cramer's V for Mulch and SiteType is: 0.0912"

#############################Visuals for the ones we want to interact
###SppGroup and powerline
TS_pow <- table(TreeLvlCombined_mod[c("SppGroup", "Powerline")])
PercTS_pow <- TS_pow/rowSums(TS_pow)
print(PercTS_pow)

#                           Powerline
# SppGroup                            0          1
#   AA_Gleditsia_triacanthos 0.59735974 0.40264026
#   Acer                     0.82517483 0.17482517
#   Acer_platanoides         0.78980892 0.21019108
#   Celtis_occidentalis      0.90410959 0.09589041
#   Fraxinus spp.            0.80625000 0.19375000
#   Other_lg                 0.75630252 0.24369748
#   Other_md                 0.70833333 0.29166667
#   Other_sm                 0.63492063 0.36507937
#   Tilia spp.               0.84210526 0.15789474
#   Ulmus spp.               0.55913978 0.44086022

###Urban App and Site Type
UA_ST <- table(TreeLvlCombined_mod[c("UrbanApp", "SiteType")])
PercUA_ST <- UA_ST/rowSums(UA_ST)
print(PercUA_ST)

#         SiteType
# UrbanApp      AA_SC      Other         SP
#        0 0.14092141 0.01355014 0.84552846
#        1 0.23829787 0.06524823 0.69645390

###Urban App and LandUse
UA_LU <- table(TreeLvlCombined_mod[c("UrbanApp", "LandUse")])
PercUA_LU <- UA_LU/rowSums(UA_LU)
print(PercUA_LU)

#         LandUse
# UrbanApp      AA_SFR     COMM_Mx        INST         MFR          MP       Other
#        0 0.517615176 0.081300813 0.006775068 0.355013550 0.017615176 0.021680217
#        1 0.411347518 0.139007092 0.039716312 0.353191489 0.026950355 0.029787234

##Urban App and Percent Imperv
UA_PI <- table(TreeLvlCombined_mod[c("UrbanApp", "PercentImperv")])
PercUA_PI <- UA_PI/rowSums(UA_PI)
print(PercUA_PI)

#         PercentImperv
# UrbanApp          1          2          3          4
#        0 0.02581522 0.05570652 0.23913043 0.67934783
#        1 0.01862464 0.02005731 0.27793696 0.68338109

##Site Type and Perc Imperv
ST_PI <- table(TreeLvlCombined_mod[c("SiteType", "PercentImperv")])
PercST_PI <- ST_PI/rowSums(ST_PI)
print(PercST_PI)

#         PercentImperv
# SiteType           1           2           3           4
#    AA_SC 0.003663004 0.003663004 0.120879121 0.871794872
#    Other 0.000000000 0.160000000 0.180000000 0.660000000
#    SP    0.027877698 0.041366906 0.294964029 0.635791367

##Site Type and Land Use
ST_LU <- table(TreeLvlCombined_mod[c("SiteType", "LandUse")])
PercST_LU <- ST_LU/rowSums(ST_LU)
print(PercST_LU)

#         LandUse
# SiteType      AA_SFR     COMM_Mx        INST         MFR          MP       Other
#    AA_SC 0.161172161 0.285714286 0.091575092 0.388278388 0.000000000 0.073260073
#    Other 0.071428571 0.000000000 0.000000000 0.785714286 0.107142857 0.035714286
#    SP    0.560538117 0.071748879 0.007174888 0.323766816 0.023318386 0.013452915

####################SpeciesGroups
OSppGroups_sm <-TreeLvlCombined_mod[TreeLvlCombined_mod$SppGroup == "Other_sm",]
totTrees_sm <- length(OSppGroups_sm$Segment)

OSppGroups_md <-TreeLvlCombined_mod[TreeLvlCombined_mod$SppGroup == "Other_md",]
totTrees_md <- length(OSppGroups_md$Segment)

OSppGroups_lg <-TreeLvlCombined_mod[TreeLvlCombined_mod$SppGroup == "Other_lg",]
totTrees_lg <- length(OSppGroups_lg$Segment)

##SppGroup - sm x UrbanApp
PercSGs_UA <- table(OSppGroups_sm$UrbanApp)/totTrees_sm
print(PercSGs_UA)

#        0        1 
# 0.953125 0.031250 

PercSGm_UA <- table(OSppGroups_md$UrbanApp)/totTrees_md
print(PercSGm_UA)

#         0         1 
# 0.4068966 0.5862069

PercSGl_UA <- table(OSppGroups_lg$UrbanApp)/totTrees_lg
print(PercSGl_UA)

#         0         1 
# 0.2833333 0.7083333 

###################Boxplots
##Basal Area x Mulch (boxplot)
ggplot(TreeLvlCombined_mod, aes(x = BasalArea, y = Mulch)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", size=2, color="red") 

#there is a range but the biggest trees have none

##Basal Area x Pruned Correct (boxplot)
TreeLvlCombined_mod$PrunedCorrect <- as.character(TreeLvlCombined_mod$PrunedCorrect )
#to be conservative here
TreeLvlCombined_mod$PrunedCorrect[is.na(TreeLvlCombined_mod$PrunedCorrect )] <- 0
TreeLvlCombined_mod$PrunedCorrect <- as.factor(TreeLvlCombined_mod$PrunedCorrect)

ggplot(TreeLvlCombined_mod, aes(x = BasalArea, y = PrunedCorrect)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", size=2, color="red")

ggplot(TreeLvlCombined_mod, aes(y = BasalArea, x = PrunedCorrect, colour = PrunedCorrect)) +
  geom_boxplot() +
  geom_jitter()

#pretty similar distributions

```

##get model runs

```{r get model runs}

#can you parallelize this?

###Set up for gjam run
TypeNames <- c("OC", "OC", "OC")

#get ydata
ydata <- TreeLvlCombined_mod[,c("Defoliation", "Discoloration", "Dieback")]

#get the model list
mlsimp <- list(ng = 20000, burnin = 5000, typeNames = TypeNames, random="TreeName")

#function
fmcomplex = ~ Segment + Year + BasalArea*SppGroup + Powerline + Mulch + Gardenscape + PrunedCorrect + CrownLight + UrbanApp*SiteType + LandUse + PercentImperv

#get x data
xdataCom <- TreeLvlCombined_mod[,c("TreeName", "Segment", "Year", "BasalArea", "SppGroup","Mulch", "Powerline", "SiteType", "Gardenscape","PrunedCorrect", "CrownLight", "UrbanApp", "LandUse", "PercentImperv")]
xdataCom$BasalArea <- log(xdataCom$BasalArea) #maybe log transform instead?

#run gjam model
out3varLG <- gjam(fmcomplex, xdataCom, ydata, modelList = mlsimp)

ppars <- list(SAVEPLOTS=T)
gjamPlot(out3var, plotPars=ppars)
saveRDS(out3var, "ChicagoTrees_3var_m1_log.rds")

###############compare DIC with version with log-transformed Basal area instead of scaled
xdataCom <- TreeLvlCombined_mod[,c("TreeName", "Segment", "Year", "BasalArea", "SppGroup","Mulch", "Powerline", "SiteType", "Gardenscape","PrunedCorrect", "CrownLight", "UrbanApp", "LandUse", "PercentImperv")]
xdataCom$BasalArea <- log(xdataCom$BasalArea)

out3var <- gjam(fmcomplex, xdataCom, ydata, modelList = mlsimp)

```


##Crown stress

```{r just street trees}
TypeNames <- c("OC", "OC", "OC")

TreeLvlCombined_CV <- TreeLvlCombined_mod[!is.na(TreeLvlCombined_mod$Gardenscape),]

##clean up the crown stress values
TreeLvlCombined_CV$CrownStress[TreeLvlCombined_CV$CrownStress == 5] <- 4
TreeLvlCombined_CV <- TreeLvlCombined_CV[!is.na(TreeLvlCombined_CV$CrownStress),]
#TreeLvlCombined_CV$CrownStress <- as.factor(TreeLvlCombined_CV$CrownStress)

#get the model list
mlsimp <- list(ng = 10000, burnin = 5000, typeNames = TypeNames, random="TreeName")

#Get the y values
TreeLvlCombined_CV$y <- TreeLvlCombined_CV$CrownStress

fmcomplexCV <- y ~ Segment + Year + BasalArea*GenusGroup + Powerline+ Mulch + Gardenscape + PrunedCorrect + UrbanApp*SiteType + LandUse

outMcomplex <- or( formula = fmcomplexCV, randVars = TreeName, data = TreeLvlCombined_CV,  ng = 10000, burnin = 1000 )



####get rid of things based on sensitivity

```


##get summary stats

need a version of the dataset that only has the most recent data for each tree

```{r summary stats}
#for supplement
TreeLvlCombined_mod$GenSpe <- TreeLvlCombined$GenSpe[match(TreeLvlCombined_mod$TreeName, TreeLvlCombined$TreeName)]
TreeLvlCombined_mod$LandUse <- TreeLvlCombined$LandUse[match(TreeLvlCombined_mod$TreeName, TreeLvlCombined$TreeName)]
TreeLvlCombined_mod$SiteType <- TreeLvlCombined$SiteType[match(TreeLvlCombined_mod$TreeName, TreeLvlCombined$TreeName)]

###always use
uniTrees <- unique(TreeLvlCombined_mod$TreeName)
colnms <- c("TreeName", "Genspe", "Genus", "SppGroup", "Segment", "BasalArea", "Discoloration", "Defoliation", "Dieback", "CrownStress", "LandUse", "PercImperv", "Powerline", "Mulch", "MulchVolcano", "CorrPrune", "Pruned", "Gardenscape", "TrunkDamage", "CrownLight", "SiteType", "DBH")
summaryDat <- as.data.frame(matrix(NA, nrow =length(uniTrees), ncol=length(colnms)))
names(summaryDat) <- colnms

for (t in 1:length(uniTrees)){
  tr <- uniTrees[t]
  
  tdat <- TreeLvlCombined_mod[TreeLvlCombined_mod$TreeName == tr,]
  mxYr <- max(as.numeric(as.character(tdat$Year)))
  mostRecDat <- tdat[tdat$Year == mxYr,]
  
  summaryDat$TreeName[t] <- tr 
  
  if (length(mostRecDat$X) > 1){
    summaryDat$Genspe[t] <- mostRecDat$GenSpe[1]
    summaryDat$Genus[t] <- mostRecDat$Genus[1]
    summaryDat$Segment[t] <- mostRecDat$Segment[1]
    summaryDat$SppGroup[t] <- mostRecDat$SppGroup[1]
    summaryDat$BasalArea[t] <- mean(mostRecDat$BasalArea)
    summaryDat$Discoloration[t] <- mean(na.omit(mostRecDat$Discoloration))
    summaryDat$Defoliation[t] <- mean(na.omit(mostRecDat$Defoliation))
    summaryDat$Dieback[t] <- mean(na.omit(mostRecDat$Dieback))
    summaryDat$CrownStress[t] <- mean(na.omit(mostRecDat$CrownStress))
    summaryDat$LandUse[t] <- mostRecDat$LandUse[1]
    summaryDat$PercImperv[t] <- mostRecDat$PercentImperv[1]
    summaryDat$Powerline[t] <- mostRecDat$Powerline[1]
    summaryDat$Mulch[t] <- mostRecDat$Mulch[1]
    summaryDat$MulchVolcano[t] <- mostRecDat$MulchVolcano[1]
    summaryDat$CorrPrune[t] <- mostRecDat$PrunedCorrect[1]
    summaryDat$Pruned[t] <- mostRecDat$Pruned[1]
    summaryDat$Gardenscape[t] <- mostRecDat$Gardenscape[1]
    summaryDat$TrunkDamage[t] <- mostRecDat$TrunkDamage[1]
    summaryDat$CrownLight[t] <- mostRecDat$CrownLight[1]
    summaryDat$SiteType[t] <- as.character(mostRecDat$SiteType[1])
    summaryDat$DBH[t] <- max(na.omit(mostRecDat$TotalDBH))
    
  } else {
    
    summaryDat$Genspe[t] <- mostRecDat$GenSpe
    summaryDat$Genus[t] <- mostRecDat$Genus
    summaryDat$Segment[t] <- mostRecDat$Segment
    summaryDat$SppGroup[t] <- mostRecDat$SppGroup
    summaryDat$BasalArea[t] <- mostRecDat$BasalArea
    summaryDat$Discoloration[t] <- mostRecDat$Discoloration
    summaryDat$Defoliation[t] <- mostRecDat$Defoliation
    summaryDat$Dieback[t] <- mostRecDat$Dieback
    summaryDat$CrownStress[t] <- mostRecDat$CrownStress
    summaryDat$LandUse[t] <- mostRecDat$LandUse
    summaryDat$PercImperv[t] <- mostRecDat$PercentImperv
    summaryDat$Powerline[t] <- mostRecDat$Powerline
    summaryDat$Mulch[t] <- mostRecDat$Mulch
    summaryDat$MulchVolcano[t] <- mostRecDat$MulchVolcano
    summaryDat$CorrPrune[t] <- mostRecDat$PrunedCorrect
    summaryDat$Pruned[t] <- mostRecDat$Pruned
    summaryDat$Gardenscape[t] <- mostRecDat$Gardenscape
    summaryDat$TrunkDamage[t] <- mostRecDat$TrunkDamage
    summaryDat$CrownLight[t] <- mostRecDat$CrownLight
    summaryDat$SiteType[t] <- as.character(mostRecDat$SiteType)
    summaryDat$DBH[t] <- mostRecDat$TotalDBH
  }
  
  
  
}

```
